---
title: "Lab 8 (Week 9)"
output: 
  html_document: 
    css: Labs.css
params:
  inc_solu:
    label: Include solutions
    value: TRUE
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

loadNamespace('haven')

data <- haven::read_sav('Lab 08 - Stress.sav')
data <- as.data.frame(data)

```

```{block include=(!params$inc_solu)}
<style>
.section.level5 {
  display: none ;
}
</style>
```

# Lab 8 - Week 9

#### Recall Lecture 8:

 - ??????????????????????????????????????????
 
### Lab Objectives

??????????????????????????????????????????

### Overview

??????????????????????????????????????????

#### Data Files

```{r echo=FALSE}
xfun::embed_file("Lab 08 - Stress.sav")
```  



## 8.1 Stress and Health

Psychologists have long been interested in the relationship between stress and health and have accumulated evidence to show that there are very real negative effects of stress on both the psychological and physical health of people.  

Researchers investigated the relationship between stress and mental health in first-year college students using a scale developed to measure the frequency, perceived importance, and desirability of recent life events. This was considered to provide a measure of a subject’s perceived social and environmental stress. Students were also asked to complete the Hopkins Symptom Check-list, assessing presence or absence of psychological symptoms.  

We wish to investigate how well the Hopkins score could be predicted based on the stress measure. The data are provided in *Lab 08 - Stress.sav*.

#### 1. Obtain a scatter plot for this data. What does this plot suggest about the relationship between stress and symptoms?

![](R.svg)

Use the `plot()` function which takes the form `plot(Y ~ X, data)`
  
![](v.svg)

Use `Analyses` ⇒ `Exploration` ⇒ `Scatterplot` (the `scatr` module must be installed)

##### Solutions

Ⓢ

The following scatterplot suggests a positive relationship between the Stress measure and the Symptoms measure, i.e., that, on average, the greater the social and environmental stress, the more symptoms of mental health problems exist.

```{r}
plot(Symptoms ~ Stress, data)
```

####  2. Fit a least squares regression model. State the equation representing the fitted least squares regression model.
  
![](R.svg)

Use the `lm()` function which takes the form `lm(Y ~ X, data)`
  
![](v.svg)

Use `Analyses` ⇒ `Regression` ⇒ `Linear Regression`  (Hint: You'll want to specify the independent variable as a `Covariate` because it is continuous)

##### Solutions

Ⓢ

$$ Symptoms = 73.890 + 0.783 \times Stress $$
  
```{r}
model <- lm(Symptoms ~ Stress, data)
model
```

#### 3. What is the coefficient of determination (R²) for the data? Interpret your result.
  
![](R.svg)

Create the model with the `lm()` function (see the previous question), and call the `summary()` function on it.
      
```{r eval=FALSE}
model <- lm(Y ~ X, data)
summary(model)
```
  
![](v.svg)

You should already have this from the previous question

##### Solutions

Ⓢ 

R² = 0.256 (see the output below) indicating that 25.6% of the variation in the Hopkins symptoms score can be explained by the linear model having the Stress measure as the predictor.

```{r}
model <- lm(Symptoms ~ Stress, data)
summary(model)
```

#### 4. Plot the residuals against the predictor variable, stress. What information does this plot give you about the model?
  
![](R.svg)

Obtain the residuals and predicted values by calling the `residuals()` and `predict()` functions on the model. you can add these values to the existing data frame and then use the `plot()` function.
      
```{r eval=FALSE}
data$residuals <- residuals(model)
data$predicted <- predict(model)
plot(residuals ~ Stress, data)
```
      
![](v.svg)

A residuals vs explanatory variable plot is available in the `Linear Regression` analysis, in the `Assumption Checks` section. Select the checkbox `Residual plots`. You'll now find a plot of the residuals against `Symptoms` in the results.

##### Solutions

Ⓢ

A plot of the residuals against stress is shown below. The existence of some non-random pattern in the residual plot suggests that there is still some information about the relationship between the response Y  (Hopkins symptoms score) and the predictor X (Stress measure) that was contained in the data but that has not been captured, or explained, by the model. Perhaps we can do better.
      
There is a suggestion of non-constant variance with the scatter of the residuals increasing for larger values of Stress.

```{r}
data$residuals <- residuals(model)
data$predicted <- predict(model)
plot(residuals ~ Stress, data)
```

#### 5. Plot the residuals against the predicted Y values. Explain how for simple linear regression this provides basically the same information as the previous plot.
  
![](R.svg)

Use the plot function, similar to the previous question
      
![](v.svg)

There's no easy way to do this in jamovi, but as the question has foreshadowed, the plot is basically the same -- only the scaling of the X-axis changes.

##### Solutions

Ⓢ 

For simple linear regression, since $\hat Y = b0 + b1X$, plotting the residuals against $\hat Y$ or against $X$ effectively provides the same graph. The only thing that changes is the scaling of the horizontal axis. (For multiple regression, the plot of the residuals against $\hat Y$ is a good plot to look at to get a quick overall summary of the residuals. However, it can still be useful on occasions to plot the residuals against one or more of the explanatory variables individually.
  
#### 6. Analyse the distribution of the residuals. Why is this important?
  
![](R.svg) perform a Q-Q plot of the residuals with the `qqPlot()` function from the `car` package:
      
```{r eval=FALSE}
library(car)
qqPlot(data$residuals)
```
      
![](v.svg) under the `Assumption Checks` section you'll find `Q-Q plot of residuals`

##### Solutions

Ⓢ

Analysis of the distribution of the residuals is shown below. The simple linear regression model assumes that the values of the response variable are normally distributed around a mean that depends linearly on the explanatory variable. The residuals are the sample estimates of this normally distributed scatter so checking the distribution of the residuals is a quick and simple check on this regression assumption. In this case the residuals appear to be right skewed, not normal. An appropriate transformation may help this, and perhaps correct the apparent non-constant variance as well.

```{r}
library(car)
hist(data$residuals, breaks=10)
qqPlot(data$residuals)
```

#### 7. We now wish to test whether a linear relationship exists between stress and symptoms. What are the null and alternative hypotheses for this test?

##### Solutions

Ⓢ

$$ H_0: β_1 = 0 $$
$$ H_a: β_1 ≠ 0 $$
      
where β₁ is the slope parameter of the line relating mean Hopkins score to the stress score for the population of interest; Hopkins score = β₀ + β₁ Stress Score.

#### 8. What is the F test statistic for this test? What is the t test statistic? What is the relationship between the t test statistic and the F test statistic for this test?
  
![](R.svg) the F test statistic is available in the summary of the model:
      
```{r eval=FALSE}
summary(model)
```
      
![](v.svg) the F test statistics is available in the `Omnibus ANOVA test`, available from the `Model Coefficients` section of the `Linear Regression` analysis.

##### Solutions

Ⓢ

Test statistic F = 36.145 and the associated p-value is 0.000 to 3 decimal places (or < 0.0005). Note that this F statistic has 1 degree of freedom in the numerator mean square. Any tests that are based on an F statistic with 1 df in the numerator mean square can also be expressed as an equivalent t-test where the t-test statistic is the square root of the F statistic. In this example we see that the test statistic t = 6.012 in the Coefficients table is the square root of F = 36.145 and the p-value is the same as that previously given.
      
```{r}
summary(model)
```

#### 9. Given an appropriate conclusion about the existence of a linear relationship between stress and symptoms.  

##### Solutions

Ⓢ

There is a statistically significant linear relationship between the stress measure and Hopkins symptoms score (p = 0.000, 3 decimal places).
      
Always quote the appropriate p-value to support your conclusion.

#### 10. Sometimes it may useful to consider confidence intervals to report results.  

  i. Obtain a 95% confidence interval for β₁ (the true slope of the regression line).  
  ii. Write an appropriate concluding sentence about this study using this confidence interval.
  iii. What advantages or disadvantages are there in using a confidence interval rather than a hypothesis test to report this information?
      
![](R.svg) confidence intervals for the parameters of the model can by obtained with the `confint()` function
      
```{r eval=FALSE}
confint(model)
```
      
![](v.svg) confidence intervals for the parameters estimates are available under the `Model Coefficients` section.

##### Solutions

Ⓢ

  i
  
```{r}
confint(model)
```

  ii
  
We are 95% confident that, on average, the Hopkins score measuring mental health symptoms increases by between 0.525 and 1.041 for every unit increase in the life events stress scale.
  
  iii
      
A confidence interval gives an estimate of the size of an effect (in this case the effect of stress due to life events) as well as an estimate of the error associated with that estimate. It can also be easily interpreted to indicate whether a particular variable is statistically significant or not at a particular significance level, α. For example: if the   (1 - α)% confidence interval for β1 does not contain 0, this is equivalent to stating that β1 is statistically different from 0 at the 5% significance level. (i.e. rejecting H₀: β1 = 0 at the 5% significance level.)
      
Since a confidence interval can indicate the size of an effect (which is about practical or clinical significance) and statistical significance (which is about the amount of evidence against a particular hypothesis), they are at times a better way of presenting results compared with a p-value alone.

#### 11. Determine a 95% confidence interval for the expected mean symptoms score for students with reported stress levels of 35?
  
![](R.svg) create a new data frame with a `Stress` value of 35, and then pass our model and this new data frame to the `predict()` function. If we provide the argument `interval='confidence'` we'll receive a confidence interval too:
      
```{r eval=FALSE}
model <- lm(Symptoms ~ Stress, data)
new_data <- data.frame(Stress=35)
predict(model, new_data, interval='confidence')
```

![](v.svg) There's no easy way to do this in jamovi, so you can copy and paste the above code into the `Rj Editor`.

##### Solutions

Ⓢ

We are 95% confident that the population mean Symptoms score for students with a reported life events stress level of 35 will be between 96.4 and 106.2 on the Hopkins scale. (Calculation using SPSS as described in the lab guide gives the same interval.) 
```{r, echo=FALSE}
model <- lm(Symptoms ~ Stress, data)
new_data <- data.frame(Stress=35)
predict(model, new_data, interval='confidence')
```

We are 95% confident that the population mean Symptoms score for students with a reported life events stress level of 35 will be between 96.4 and
106.2 on the Hopkins scale.

#### 12. Calculate a 95% prediction interval for the symptoms score of a student who reported a stress level of 35? Also compare this result with the 95% confidence interval.

![](R.svg) Same as the call to `predict()` in the previous question but with `interval='prediction'`.

```{r eval=FALSE}
model <- lm(Symptoms ~ Stress, data)
new_data <- data.frame(Stress=35)
predict(model, new_data, interval='prediction')
```

![](v.svg) Copy and paste the above code into the `Rj Editor`.

##### Solutions

Ⓢ

```{r, echo=FALSE}
model <- lm(Symptoms ~ Stress, data)
new_data <- data.frame(Stress=35)
predict(model, new_data, interval='prediction')
```

We are 95% confident that the symptoms score for an individual student with a reported life events stress level of 35 will be between 66.1 and 136.5 on the Hopkins scale. The statements given above identify the differences between these two intervals and to what they refer. The first interval is an estimate of the mean.

Hopkins symptoms score for the population of students who score $X_h$ = 35 on the stress measure. The second interval is an estimate of where the Hopkins score for any one of those individuals might be. Notice the increased width of the latter.
